---
title: "One-Way ANOVA vignette"
author: "Author: Antdrew D. Nguyen"
date: "Date: 2017-06-01"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    toc: TRUE
  html_notebook: default
---

# Libraries

```{r}
library(plyr)
library(tidyr)
```

# Reviewing Analysis of Variance (ANOVA) 


I'll be following Chapter 10 of *A Primer of Ecological Statistics* by Gotelli and Ellison  

Key Concepts: 

ANOVA aims to determine differences in a continuous variable between 2 or more groups.   

ANOVA bulit on partitioning on the concept of paritioning of the sum of squares ($SS_{total}$). How do we calculate this? 

The total sum of squares of the data is the sum of squared deviations of each observation ($Y_i$) from the grand mean ($\bar{Y}$). There are $i$ = 1 to $a$ treatment levels and $j$ = 1 to $n$ replicates per treatment. 

**recap**:  

* $i$ refers to the treatment levels   
    * it looks like $a$ represents the number of different treatments   
* $j$ refers to each observations, with $n$ being the number of replicates    


$$ SS_{total} = \sum_{i=1}^a\sum_{j=1}^n (Y_{ij}-\bar{Y})^2$$

The total sum of squares is the deviation of each observation from the grand mean. 

$SS_{total}$ can then be partitioned into different components, mainly **among and within groups**.  

$$SS_{total} = SS_{among} + SS_{within}$$
So for the *among* group SS: 

$$ SS_{among} = \sum_{i=1}^a\sum_{j=1}^n (\bar{Y}_{i}-\bar{Y})^2$$
So for the *within* group SS: 

$$ SS_{within} = \sum_{i=1}^a\sum_{j=1}^n (Y_{ij}-\bar{Y}_i)^2$$

**Bringing it all together**:    

$$ \sum_{i=1}^a\sum_{j=1}^n (Y_{ij}-\bar{Y})^2 =  \sum_{i=1}^a\sum_{j=1}^n (\bar{Y}_{i}-\bar{Y})^2 +  \sum_{i=1}^a\sum_{j=1}^n (Y_{ij}-\bar{Y}_i)^2$$


## Enter data   

```{r}
#data, 
# a = 3, 3 treatments
# n = 4, 4 reps
n=4
unman<-c(10,12,12,13)
control<-c(9,11,11,12)
treat<-c(12,13,15,16)

wide.dat<-data.frame(unman,control,treat);wide.dat
long.dat<-gather(wide.dat,treatment,measure,unman:treat);long.dat
#global mean
grandmean<-round(mean(c(unman,control,treat)),2);grandmean
```

## $SS_{total}$ calculations   

```{r}
sum((long.dat$measure-grandmean)^2)
```


## $SS_{among}$ calculations  

```{r}
## calculating 1 case

(mean(unman)-grandmean)^2

### making a whole function 
#with ddply
ssa<-function(n=n,vec=c(1,3,3),grandmean=grandmean){
  SSa<-(mean(vec)-grandmean)^2
  SSa
}
ssa(vec=unman,n=n,grandmean=grandmean) # verify function

## executing function
ssam<-ddply(long.dat,.(treatment),summarize,ssamong=ssa(vec=measure,n=n,grandmean=grandmean));ssam

SSAM<-n*sum(ssam$ssamong);SSAM
# for a balanced design!
```


## $SS_{within}$ calculations  

```{r}

## calculating 1 case
sum((unman-mean(unman))^2)


### making a whole function 
#with ddply
sswi<-function(x){
  SSwithin<-sum((x-mean(x))^2)
  SSwithin
}
sswi(unman) # verify function

SSwi<-ddply(long.dat,.(treatment),summarize,sswi=sswi(measure));SSwi

sumwithin<-sum(SSwi$sswi);sumwithin
```



# Assumptions of ANOVAs

1. Samples are indepednent and identically distributed. 
2. Variances are homogeneous among groups   
    * variance within each group approx to variance within other groups    
3. Residuals are normally distributed  
4. Samples are classified correctly   
5. Maine effects are additive  

# Hypothesis testing   

Definitions: 

* $Y_{ij}$: replicated $j$ associated with treatment level $i$  
* $\mu$ is the *true* grand mean or average   
* $A_i$ is the additive linear component associated with level $i$ of treatment $A$. 
  * There is a different coefficient $A_i$ associated with each treatment level ($i$).  
  * positive coefficients mean that the treatment level has a higher value than grand mean  


Alternative Hypothesis, $H_a$: $Y_{ij} = \mu + A_i + \epsilon_{ij}$

Null Hypothesis, $H_o$: $Y_{ij} = \mu + \epsilon_{ij}$

# Verify with aov() function 

```{r}
knitr::kable(round(summary(aov(measure~treatment,data=long.dat))[[1]],2))

```
