---
title: "2018-01-10_simulate_perf_curves"
author: "Antdrew D. Nguyen, Evolutionary Physiologist"

output: 
  html_document: 
    theme: cerulean
    toc: yes
---

# Libaries
```{r, include=FALSE}
library(ggplot2)
library(lmerTest)
library(lme4)
library(DiagrammeR)
library(gridExtra)

### figure set up 

##figure set up 
setup<-theme_bw()+theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),legend.position="none",
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),panel.background = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank())

```

# Workflow

Simulate 3 possible outcomes for how performance curves (modeled as a gaussian function) can vary: 

1. Veritical --traits are all correlated across environmental gradient
2. Horizontal -- optimum varies along environmental gradient; so genotypes/clones have different optimum
3. Generalist-specialist -- those with higher max performance, also operate over a smaller window along the environmental gradient 


For each possibility:  
```{r}
mermaid("
graph TD
  a{Simulate Data} --20 clones or colonies, 5 reps--> b(Mixed Effect Model) 
  b --> c[Extract G] 
  c --> d[PCA decomposition] 
  d --> e[Visualize]
  a --> e
")
```

# Constructing gaussian shaped performance curves  

$$Performance = height * e^{(-0.5(\frac{Temperature - T_{opt}}{width})^2)}$$

```{r gaussian curve construction}

gau<-function(a=1,b=30,c=10,t=seq(1,60,1)){
  #z=a*exp(-.5((t-b)/c)^2)
  z=a*exp(-.5*((t-b)/c)^2)
  b<-data.frame(performance=z,t=t)
  return(b)  
}

#gau()
plot(gau()$t,gau()$performance,type="l")
```


## Simulate data with just vertical variation/responses

### making the data

```{r}
#simulate vert values
vval<-rnorm(mean=50,sd=15,n=100)

#pull out 20, randomly

col<-sort(rep(sample(vval,20),5))


#estimate perf curves
random.error<-sample(rnorm(mean=0,sd=.5,n=1000),500)

#generate dataset with 5 reps for 20 colonies for 4 levels of temperature
dat<-rbind(gau(a=col,t=c(20,25,30,35,40)),gau(a=col,t=c(20,25,30,35,40)),gau(a=col,t=c(20,25,30,35,40)),gau(a=col,t=c(20,25,30,35,40)),gau(a=col,t=c(20,25,30,35,40)))
dat$performance<-dat$performance+random.error


#convert colonies and dat into factors
dat$colonies<-as.factor(rep(sort(rep(seq(1,20,1),5)),5))
dat$t<-as.factor(dat$t)

```

### Mixed effects model to estimate G

```{r}
### building mixed effects model to estimate the variance-covariance structure
# random slope and temperature nested within colonies
vmod<-lmer(formula=performance~1+(0+t|colonies),REML=TRUE,data=dat)
#summary(vmod)
Gmatrix.pac<-VarCorr(vmod)$colonies;Gmatrix.pac
Gmatrix.pac[1:5,1:5]
```

### PCA decomposition of G  

```{r}
h<-princomp(Gmatrix.pac[1:5,1:5])
h$loadings[,1:3]
h1<-data.frame(Loadings=h$loadings[,1],t=c(20,25,30,35,40))

aa<-ggplot(h1,aes(x=t,y=Loadings))+geom_line(size=3)+setup+geom_hline(yintercept=0)+ylim(-1,1)
aa
```

* y axis = loadings 
* x axis = environmental gradient

### Figure
```{r vertical variation, warning=FALSE}
dat$t<-as.numeric(as.character(dat$t))
a<-ggplot(dat,aes(x=t,y=performance,colour=factor(colonies)))+geom_smooth(se=FALSE)+setup+scale_color_hue(l=20, c=50)+ylim(5,100)
a

```

## Simulate horizontal variation

```{r horizontal variation}
#simulate Topt values
opt<-rnorm(mean=30,sd=5,n=100)

#pull out 20, randomly
col.opt<-sort(rep(sample(opt,20),5))
col.opt

#estimate perf curves
#random.error<-sample(rnorm(mean=0,sd=.5,n=1000),500)

#generate dataset with 5 reps for 20 colonies for 4 levels of temperature
dat2<-rbind(gau(b=col.opt,t=c(20,25,30,35,40)),gau(b=col.opt,t=c(20,25,30,35,40)),gau(b=col.opt,t=c(20,25,30,35,40)),gau(b=col.opt,t=c(20,25,30,35,40)),gau(b=col.opt,t=c(20,25,30,35,40)))
dat2$performance<-dat2$performance#+random.error

###
dat2$colonies<-as.factor(rep(sort(rep(seq(1,20,1),5)),5))
dat2$t<-as.factor(dat2$t)

```

### Mixed effects model to Estimate G
```{r}
### building mixed effects model to estimate the variance-covariance structure
# random slope and temperature nested within colonies
hmod<-lmer(formula=performance~1+(0+t|colonies),REML=TRUE,data=dat2)
#summary(hmod)
Gmatrix.pac<-VarCorr(hmod)$colonies;Gmatrix.pac
Gmatrix.pac[1:5,1:5]
```


### PCA decomposition of G

```{r}
h<-princomp(Gmatrix.pac[1:5,1:5])
h$loadings[,1:3]

h2<-data.frame(Loadings=h$loadings[,1],t=c(20,25,30,35,40))

bb<-ggplot(h2,aes(x=t,y=Loadings))+geom_line(size=3)+setup+geom_hline(yintercept=0)+ylim(-1,1)
bb
```

* y axis = loadings 
* x axis = environmental gradient

### Figure

```{r horizontal variation figure, warning=FALSE}
dat2$t<-as.numeric(as.character(dat2$t))
b<-ggplot(dat2,aes(x=t,y=performance,colour=factor(colonies)))+stat_smooth(se=FALSE)+setup+scale_color_hue(l=20, c=50)+xlim(20,40)+ylim(0.5,1.5)
b
```

## Simulate generalist-specialist variation

### making the data

ones that have small widths, have higher performance   

```{r gen specialist data}
col# height

w<-rnorm(mean=5,sd=3,n=100)
col.opt<-sort(rep(sample(w,20),5))
col.opt # width

dat3<-rbind(gau(a=col,c=rev(col.opt),t=c(20,25,30,35,40)),gau(a=col,c=rev(col.opt),t=c(20,25,30,35,40)),gau(a=col,c=rev(col.opt),t=c(20,25,30,35,40)),gau(a=col,c=rev(col.opt),t=c(20,25,30,35,40)),gau(a=col,c=rev(col.opt),t=c(20,25,30,35,40)))
#dat3
dat3$colonies<-as.factor(rep(sort(rep(seq(1,20,1),5)),5))
dat3$t<-as.factor(dat2$t)
```

### Mixed effects model to estimate G

```{r mixed efefcts model gen specialist}
###model
hwmod<-lmer(formula=performance~1+(0+t|colonies),REML=TRUE,data=dat3)
#summary(hmod)
Gmatrix.pac<-VarCorr(hwmod)$colonies;Gmatrix.pac
Gmatrix.pac[1:5,1:5]
```


### PCA decomposition of G   

```{r pca decomp gen speci}

h<-princomp(Gmatrix.pac[1:5,1:5])
h$loadings[,1:3]
summary(h)
#plot(h$loadings[,1])
h3<-data.frame(Loadings=h$loadings[,1],t=c(20,25,30,35,40))

cc<-ggplot(h3,aes(x=t,y=Loadings))+geom_line(size=3)+setup+geom_hline(yintercept=0)+ylim(-1,1)
cc
```

* y axis = loadings 
* x axis = environmental gradient

### Figure

```{r gen-spec variation figure, warning=FALSE}
dat3$t<-as.numeric(as.character(dat3$t))
c<-ggplot(dat3,aes(x=t,y=performance,colour=factor(colonies)))+stat_smooth(se=FALSE)+setup+scale_color_hue(l=20, c=50)
c
```

# Summary, all figs together  

```{r warning=FALSE}
grid.arrange(ncol=2,nrow=3,a,aa,b,bb,c,cc)
```

# References: 

* **using mixed effects models, I followed [Paccard et al. 2016](http://www.journals.uchicago.edu/doi/abs/10.1086/685643):**   

Paccard, A., Van Buskirk, J., & Willi, Y. (2016). Quantitative Genetic Architecture at Latitudinal Range Boundaries: Reduced Variation but Higher Trait Independence. The American Naturalist, 187(5), 667–77. https://doi.org/10.1086/685643


* **The idea of decomposing the population level variation (with known kinship) of performance curves , I followed [Joel Kingsolver's work](http://jgking.web.unc.edu/publications/):** 

Kingsolver, J.G., N. Heckman, J. Zhang, P.A. Carter, J.L. Knies, J.R. Stinchcombe, and K. Meyer. 2015. Genetic variation, simplicity, and evolutionary constraints for function-valued traits. The American Naturalist 185: E166-E181.


Stinchcombe, J. R., Function-valued Traits Working Group, M. Kirkpatrick. 2012. Genetics and evolution of function-valued traits:  understanding environmentally responsive phenotypes.  Trends in Ecology and Evolution.   


Kingsolver, J. G., G. J. Ragland, and J. G. Shlichta. 2004. Quantitative genetics of continuous reaction norms: Thermal sensitivity of caterpillar growth rates. Evolution 58:1521-1529.





# SessionInfo

```{r Sessioninfo}
sessionInfo()
```