---
title: "PCA of rasters"
author: "Antdrew D. Nguyen"
date: "2016-12-07"
output: 
  
  prettydoc::html_pretty:
    theme: cayman
    toc: yes
    toc_depth: 2
---
<style type="text/css">

body{ /* Normal  */
   font-size: 24px;
}

code.r{ /* Code block */
  font-size: 18px;
}
pre { /* Code block */
  font-size: 18px
}
</style>

This is a data demo for performing a PCA on multiple climate raster layers with [*rasterPCA()* function](http://bleutner.github.io/RStoolbox/rstbx-docu/rasterPCA.html).


# Load libraries and grabbing dataset
```{r load libraries}
#libraries related to maps
library(sp)
library(raster)
library(maps)
library(mapdata)
library(RStoolbox)# rasterPCA function

#packages for reading in data
library(data.table)


#reading in dataset
dat<-fread("https://raw.githubusercontent.com/adnguyen/HelmsCahan_CBP-partA_2016/master/Script_Analyses/Sampling_sites_table.csv")

```

## looking at the data
```{r}
dat<-data.frame(dat)
str(dat)
```

## grab bioclim climate variables and add to dataset
```{r grab bioclim var and dataset}
w <- getData('worldclim', var='bio', res=2.5) # downloading data
plot(w,1) # plot of mean annual temperature

#dbio1 <- extract(w, dat[,c("Longitude","Latitude")]) # by column
dbio1 <- extract(w, dat[,5:6]) #grabbing bioclim variables based on coordinates
full.dat <- data.frame(cbind(dat, dbio1[,1:19])) # combining orginal data set with bioclim variables

```

# Plot eastern US unmanipulated climate data
```{r}

plot(w, 1, xlim=c(-130, -65), ylim=c(25,50), axes=F, legend=F, col=colorRampPalette(c("lightblue","orange"))(255),main="",box=FALSE)
map("worldHires",c("USA","Canada"),add=TRUE)
map("state", c('florida', 'south carolina', 'north carolina', 'georgia', 'virginia', 'west virginia', 'maryland', 'delaware', 'new jersey', 'rhode island', 'new york', 'connecticut', 'massachusetts', 'pennyslvania', 'vermont', 'new hampshire', 'maine', 'alabama', 'tennessee', 'kentucky', 'ohio','iowa','illinois','arkansas','missouri','minnesota','wisconsin','michigan','louisiana','mississippi',"texas","arizona","illinois","california","oregon","utah","washington","kansas","new mexico","montana","idaho","wyoming","north dakota","south dakota","nebraska","oklahoma"), add = TRUE)
```

# Using rasterPCA function   
Climate is usually correlated and there are a lot of them. One way to reduce the complexity is to perform a PCA on those variables (19 of them). Also, PCA'ing the whole word will take a long time, so crop the desired region first. 
```{r rasterPCA}
#look at range of lat and lon
range(full.dat[,5]) #lon range
range(full.dat[,6]) # lat range
#designating limits
lims<-c(-85,-65,30,50)

submap<-crop(w,lims)
plot(submap,1)#plot MAT
map("worldHires",c("USA","Canada"),add=TRUE)
map("state", c('florida', 'south carolina', 'north carolina', 'georgia', 'virginia', 'west virginia', 'maryland', 'delaware', 'new jersey', 'rhode island', 'new york', 'connecticut', 'massachusetts', 'pennyslvania', 'vermont', 'new hampshire', 'maine', 'alabama', 'tennessee', 'kentucky', 'ohio','iowa','illinois','arkansas','missouri','minnesota','wisconsin','michigan','louisiana','mississippi',"texas","arizona","illinois","california","oregon","utah","washington","kansas","new mexico","montana","idaho","wyoming","north dakota","south dakota","nebraska","oklahoma"), add = TRUE)


#now we can do pca
pcamap<-rasterPCA(submap,spca=TRUE)
#check loadings and eigenvalues
knitr::kable(round(pcamap$model$loadings[,1:3],3)) # top 3 loadings
#eigenvalues
summary(pcamap$model)

#plot first pca1
plot(pcamap$map,1) #plot the first pca!
map("worldHires",c("USA","Canada"),add=TRUE)
map("state", c('florida', 'south carolina', 'north carolina', 'georgia', 'virginia', 'west virginia', 'maryland', 'delaware', 'new jersey', 'rhode island', 'new york', 'connecticut', 'massachusetts', 'pennyslvania', 'vermont', 'new hampshire', 'maine', 'alabama', 'tennessee', 'kentucky', 'ohio','iowa','illinois','arkansas','missouri','minnesota','wisconsin','michigan','louisiana','mississippi',"texas","arizona","illinois","california","oregon","utah","washington","kansas","new mexico","montana","idaho","wyoming","north dakota","south dakota","nebraska","oklahoma"), add = TRUE)


```

## compare pca of climate of sites vs raster layer
```{r compar pcas}
datpca<-princomp(scale(full.dat[,8:26]))
knitr::kable(round(datpca$loadings[,1:3],3))# pca of sites

#looks similar
```



# Session Information
```{r session information}
sessionInfo()
```

